---
title: One Health VBD Hub - R Package
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

<!-- Build with devtools::build_readme() -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<!-- # One Health VBD Hub - R Package -->

<!-- badges: start -->
```{r, echo = FALSE}
dep <- as.vector(read.dcf('DESCRIPTION')[, 'Depends'])
m <- regexpr('\\d\\.*\\d+\\.*\\d*\\.*\\d*', dep)
rm <- regmatches(dep, m)
rvers <- rm
# rvers <- gsub('.*(\\\\d+.\\\\d+.\\\\d+).*', '\\\\1', rm)
# rm
```

```{r, echo = FALSE}
version <- as.vector(read.dcf('DESCRIPTION')[, 'Version'])
version <- gsub('-', '.', version)
```

[![R](https://img.shields.io/badge/R%3E%3D-`r rvers`-6666ff.svg?style=for-the-badge)](https://cran.r-project.org/)
[![packageversion](https://img.shields.io/badge/Package%20version-`r version`-orange.svg?style=for-the-badge)](commits/master)
[![license](https://img.shields.io/badge/license-GPL--3-blue.svg?style=for-the-badge)](https://www.gnu.org/licenses/gpl-3.0.en.html)

[![R-CMD-check](https://github.com/fwimp/ohvbd/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/fwimp/ohvbd/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->

## Introduction

`ohvbd` is an R package for retrieving (and parsing) data from a network of disease vector data sources.

At present this package is very much in its early stages, as such the internal API may change without warning.

## Installation

You can install the development version of ohvbd from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("fwimp/ohvbd", build_vignettes = TRUE)
```

