---
title: "Retrieving data from VectorByte"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Retrieving data from VectorByte}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

## Introduction

The ohvbd package allows you to retrieve data from many different databases directly.

Currently these databases include the [VecTraits](https://vectorbyte.crc.nd.edu/vectraits-explorer) and [VecDyn](https://vectorbyte.crc.nd.edu/vecdyn-datasets) projects from [VectorByte](https://www.vectorbyte.org/).

Let's imagine you wanted to figure out where a particular vector species - _Aedes aegypti_, for example - was found. Here's how you'd use `ohvbd` to do that:

```{r exemplar, echo=TRUE, eval=FALSE}
df <- search_vt("Aedes aegypti") %>%
  get_vt() %>%
  extract_vt(
    cols = c("Interactor1Genus", "Interactor1Species", "Latitude", "Longitude"),
    returnunique = TRUE
  )
df
```

Great, lovely. But now... what does that all mean? And how can you build up your own data request? Read on to find out...

## Some notes before we begin

If you are using Linux, sometimes `httr2` fails to communicate with the VectorByte servers. As such on these systems it is best to set ohvbd to use slightly less stringent identity checks in its calls to the server.

You only need to perform this once per R session / script.
```{r earlypkgload, echo=FALSE}
library(ohvbd)
```
<!-- Shim to make sure vignette builds on ubuntu-->
```{r option, echo=TRUE, message=FALSE, results=FALSE}
get_os <- function() {
  sysinf <- Sys.info()
  if (!is.null(sysinf)) {
    os <- sysinf["sysname"]
    if (os == "Darwin")
      os <- "osx"
  } else { ## mystery machine
    os <- .Platform$OS.type
    if (grepl("^darwin", R.version$os))
      os <- "osx"
    if (grepl("linux-gnu", R.version$os))
      os <- "linux"
  }
  tolower(os)
}
if (get_os() == "linux") {
  set_ohvbd_compat()
}
```

For users who are familiar with tidyverse pipes (`%>%`), this approach is generally usable in ohvbd as well.

For users who are not familiar, pipes take the output of one command and feed it forward to the next command as the first argument:


```{r nonpipedemo}
# Find mean of a vector normally
x <- c(1, 2, 3)
mean(x)
```

```{r pipedemo}
# Find mean of a vector using pipes
c(1, 2, 3) %>% mean()
```

For the rest of this vignette we will be using a piped-style approach.

Another note before we begin: the functions used to interface with VecTraits and VecDyn are basically the same, except for a `vt` or `vd` in the function name. So `search_vt()` is the VecTraits equivalent of `search_vd()`. We will use VecTraits functions in this vignette, but the process is the same for VecDyn.

## Finding IDs

Datasets in VecTraits and VecDyn are organised by id. As such it is often good to be able to see which IDs are currently available. We can do this using the `get_current_x_ids()` family of functions. These return a vector of valid IDs on the given database:

```{r get_vt_ids}
library(ohvbd)
get_vt_current_ids() %>% head(n = 20)
```

The functions `search_x()` and `search_x_smart()` also return a list of IDs that match the query provided in the arguments.

In this case let's search VecTraits for *Aedes aegypti*, the "Yellow Fever mosquito":

```{r search_vt}
aedes_ids <- search_vt("Aedes aegypti")
aedes_ids
```

This again returns a vector of IDs, this time corresponding to any dataset with *Aedes aegypti* in one of the searchable columns.

You can also find IDs interactively online by using the [VecTraits Explorer](https://vectorbyte.crc.nd.edu/vectraits-explorer) should you so choose.

## Getting data

Now you have a vector of datasets, we need to actually retrieve the data of these datasets through the API.

To do this we can use the `get_x()` functions. In this case let's get the first 5 *Aedes aegypti* datasets:

```{r get_vt, echo=TRUE, message=FALSE, results=FALSE}
aedes_responses <- get_vt(aedes_ids[1:5])
aedes_responses[[1]]
```
```
## <httr2_response>
## GET https://vectorbyte.crc.nd.edu/portal/api/vectraits-dataset/97/?format=json
## Status: 200 OK
## Content-Type: application/json
## Body: In memory (26362 bytes)
```

The `get_vt()` function returns a list of the data in the form of the original `httr2` responses. These are useful if you want to know specifics about how the server sent data back, but for most usecases it is more useful to extract the data into a dataframe.

## Extracting data

Now we have a list of responses, we can extract the relevant data from them using the `extract_x()` functions.

```{r extract_all_vt}
aedes_data <- aedes_responses %>% extract_vt()
paste0("Data dimensions: ", ncol(aedes_data), " cols x ", nrow(aedes_data), " rows")
```

This dataset is a bit too large to print here, and often you may only want a few columns of data rather than the whole dataset.

Fortunately the `cols` argument allows us to filter this data out easily. We can also use the argument `returnunique` to instruct ohvbd to return only unique rows.

So let's get just the unique locations and trait name combinations from our data using the same command as before:

```{r extract_filtered_vt}
aedes_data_filtered <- aedes_responses %>%
  extract_vt(
    cols = c("LocationText", "OriginalTraitName"),
    returnunique = TRUE
  )
aedes_data_filtered
```

## Chunked downloading

In some cases you may end up downloading a lot of data. This can cause problems if you do not have enough memory on your computer to store this data in memory all at once, even if you just want one or two columns, like before,

One potential method for mitigating this is to use the `get_extract_x_chunked()` functions. These only download a few records at a time and extract them before moving on to the next chunk of ids. We can perform the filtered extraction from above like so:

```{r get_extract_vt_chunked}
get_extract_vt_chunked(
  aedes_ids[1:5],
  chunksize = 2,
  cols = c("LocationText", "OriginalTraitName"),
  returnunique = TRUE
)
```
## Putting it all together

In day-to-day use, you will mostly find yourself using all these functions together to create small pipelines.

A typical pipeline would likely only contain a few lines of code:

```{r vt_pipeline, eval=FALSE, echo=TRUE, message=FALSE, results=FALSE}
df <- search_vt("Aedes aegypti") %>%
  get_vt() %>%
  extract_vt(
    cols = c("Interactor1Genus", "Interactor1Species", "Latitude", "Longitude"),
    returnunique = TRUE
  )
head(df)
```
```{r vt_pipeline_secret, echo=FALSE, message=FALSE}
ids <- search_vt("Aedes aegypti")
df <- get_vt(ids[1:5], check_src = FALSE) %>%
  extract_vt(
    cols = c("Interactor1Genus", "Interactor1Species", "Latitude", "Longitude"),
    returnunique = TRUE
  )
head(df)
```


## Smart searching

One subtlety of the VectorByte search interfaces is to do with name collisions.

Let's imagine that we are looking for traits of whitefly species (*Bemisia*). We can just construct a query to investigate this as follows:


```{r demo_searcherror}
df <- search_vt("Bemisia")[1:6] %>%
  get_vt(check_src = FALSE) %>%
  extract_vt(cols = c("DatasetID", "Interactor1Genus", "Interactor1Species", "Interactor2Genus", "Interactor2Species"))
```

Now we would expect this to be traits of *Bemisia* spp, however when we look at the Interactor1Genus column we see something a touch odd:

```{r}
unique(df$Interactor1Genus)
```

*Axinoscymnus* is a ladybird, but why is it appearing? Let's look at only rows containing *Axinoscymnus*:

```{r}
df %>% dplyr::filter(Interactor1Genus == "Axinoscymnus") %>% head()
```

In this scenario, *Bemisia* is present in the dataset, but it is as the "target" rather than the animal that the trait is referring to.

As such we might want to be more specific about precisely which data to retrieve. Enter the `search_x_smart()` family of functions.

These allow you to construct a more specific search.

So let's construct the same search as we were wanting to do before, but with the smart search.

```{r}
df <- search_vt_smart("Interactor1Genus", "contains", "Bemisia")[1:6] %>%
  get_vt(check_src = FALSE) %>%
  extract_vt(cols = c("DatasetID", "Interactor1Genus", "Interactor1Species", "Interactor2Genus", "Interactor2Species"))
unique(df$Interactor1Genus)
```

Here we have made sure to only search the `Interactor1Genus` column. As such we have only gotten back *Bemisia* traits!

This same sort of collision is particularly common in the "Citation" column, where papers may mention multiple trait names.

The `search_x_smart()` functions have many different operators and columns that they can work upon. For full details, run `?search_vt_smart` or `?search_vd_smart` in your console.

In general it is always worthwhile inspecting the data you retrieve from VectorByte to make sure that your query returned the data that you thought it did.

## Futher steps

From here you now have all the data you might need for further analysis, so now it's down to you!

One final note to end on: it can be advisable to save any output data in a csv or parquet format so that you do not need to re-download it every time you run your script. This is as easy as running `write.csv()` on your dataframe, then reading it in later with `read.csv()`
